Real-space computation of $E$/$B$-mode maps I: \revisit{Formalism, Compact Kernels, and Polarized Filaments}

### 2018-06-21 articles/formalism/article.tex
Edit on title and discussion section.

\begin{document}
%\title{Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$}
%\title{Real space operators for Stokes $Q$ \& $U$ to scalars $E$ \& $B$ translation}
\title{Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels}
\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, 77 Chieftan Way, Tallahassee, Florida, U.S.A.}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

\abstract{
\noindent 
\revisit{We derive full-sky, real-space operators that convert between polarization Stokes $Q/U$ parameters to the coordinate-independent scalar $E$/$B$ modes that are widely used in Cosmic Microwave Background(CMB) and cosmic shear analysis. We also derive real space operators that decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations.  For all these real space operators we show that the kernels split naturally into angular and radial parts and we show explicitly how the radial extent of these kernels depends on the targeted band-limit. We show that the kernels can be interpreted either as a complex convolving beam or the Green's function when they are expressed as functions of the forward or inverse rotation Euler angles respectively. We show that an arbitrary radial dependence, satisfying the constraint that it vanishes at angular separations of $0$ and $\pi$, can produce $E/B$-like maps and that these are simply filtered versions of the standard $E/B$ maps.  This allows us to compute $E/B$ maps in real space with a compactly-supported kernel, an approach that can guarantee the avoidance of known foreground regions and can be employed in a massively parallel scheme at high-resolution. We show that the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators are a special case of these generalized operators and present their band limited versions. We can recover the standard power spectrum of the polarized CMB sky by correcting the power spectra of these maps with a simple window function. We present a scheme to derive this window function for any radial dependence. The spatial structure of the real space operator provides great intuition for the $E/B$ structure of polarized, filamentary galactic foregrounds and we predict non-zero B-mode signature that is expected from finite filaments in the sky. This paper is the first part in a series of papers that explore real-space computation of polarization modes and their applications.
  } 
  \\
  \\
%  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. \revisit{This is part 1 of a series of papers where we have derived the real space operators. In the following papers we will explore some of their applications.} \comment{Needs better blending.}
}

>>>>

\begin{document}
%\title{Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$}
%\title{Real space operators for Stokes $Q$ \& $U$ to scalars $E$ \& $B$ translation}
%\title{Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels}
\title{Real-space computation of $E$/$B$-mode maps I: \revisit{Formalism, Compact Kernels, and Polarized Filaments}}

\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, 77 Chieftan Way, Tallahassee, Florida, U.S.A.}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

\abstract{
\noindent 
\revisit{We derive full-sky, real-space operators that convert between polarization Stokes $Q/U$ parameters to the coordinate-independent scalar $E$/$B$ modes that are widely used in Cosmic Microwave Background(CMB) and cosmic shear analysis. We also derive real space operators that decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations.  For all these real space operators we show that the kernels split naturally into angular and radial parts and we show explicitly how the radial extent of these kernels depends on the targeted band-limit. We show that the kernels can be interpreted either as a complex convolving beam or the Green's function when they are expressed as functions of the forward or inverse rotation Euler angles respectively. We show that an arbitrary radial dependence, satisfying the constraint that it vanishes at angular separations of $0$ and $\pi$, can produce $E/B$-like maps and that these are simply filtered versions of the standard $E/B$ maps.  This allows us to compute $E/B$ maps in real space with a compactly-supported kernel, an approach that can guarantee the avoidance of known foreground regions and can be employed in a massively parallel scheme at high-resolution. We show that the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators are a special case of these generalized operators and present their band limited versions. We can recover the standard power spectrum of the polarized CMB sky by correcting the power spectra of these maps with a simple window function. We present a scheme to derive this window function for any radial dependence. The spatial structure of the real space operator provides great intuition for the $E/B$ structure of polarized, filamentary galactic foregrounds and we predict a non-zero $B$-mode signature that is expected from finite filaments in the sky. This paper is the first part in a series of papers that explore real-space computation of polarization modes and their applications.
  } 
  \\
  \\
%  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. \revisit{This is part 1 of a series of papers where we have derived the real space operators. In the following papers we will explore some of their applications.} \comment{Needs better blending.}
}

### 2018-06-21 articles/formalism/sections/discussion.tex
Edit on title and discussion section.

\section{Discussion}\label{sec:discussion}

We introduced a vector-matrix notation which allows for concise book keeping of all the standard operations involved in the analysis of CMB polarization maps. This notation simplifies the derivation of the real space operators. We presented a first derivation of the real space operators on the sphere that transforms the Stokes vector \vp{} to into a vector of scalars \vs{} and vice versa. We also presented real space operators that directly decompose the full Stokes vector \vp{} into vector \vp{E} and \vp{B} that correspond to the respective scalar modes.

These real space operators provide a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators have the common characteristic  of being separable into a azimuthal part that is band limit independent and radial part that incorporates all the band limit dependence. The azimuthal part of the operator is primarily responsible for the requisite spinorial decomposition, while the radial weights determine the non-local dependence of the construction of the resultant fields on the original fields. Using the self similarity property of the radial functions we define a non-locality parameter and show that it is proportional to $\ell_{\rm max}^{-1}$. The scalar decomposition of the Stokes parameters being of primary interest,  we empirically derived the non-locality parameter $\beta_0$ for this operation and showed that the proportionality constant $\ell_{0}=22$ provided a good prediction for the angular distance at which the ${}_{\mm}f$ falls below 1\% of its maxima as seen in \fig{fig:rad_ker_decay}.

Our careful study of the real space operators reveals the dual interpretation of the operators as either the Green's function or a complex convolving beam depending on whether the kernels are expressed in terms of the forward or inverse rotation Euler angles. While the convolution interpretation is a familiar one, the Green's function interpretation is a new way of looking at this operation. This new interpretation allows us to think of ${}_{+2}X$ as some spin-2 charge which radiates out a complex spin-0 scalar field $E+iB$. The resultant complex scalar maps can be then understood as arising from superposition of the radiating field emanating from all the spin charges on the sphere. The $E/B$ mode maps are merely the real and imaginary parts of this resultant complex scalar field.

\revisit{A similar equation for real space $E$ \& $B$ operators was derived in \cite{Zaldarriaga2001a}, however those results were derived for the flat sky case and did not explicitly derive the radial kernel.} \comment{A discussion on this should be in the conclusions.}

Understanding the non-locality of the real space operators, allowed us to generalize the real space operators $\bar{O}'$ and its inverse ${\bar O}'^{-1}$, which transform between the spin-2 and spin-0 representations of the CMB polarization. We presented a method of systematically generalizing these operators by introducing the harmonic space operator $\tilde{\mathcal{G}}$ which in effect causes the radial function to vary from its default form. We show that these modifications to the radial kernel can be interpreted as a smoothing operation on the scalar fields with a circularly symmetric instrument beam. We argue that one can choose any arbitrary form for $\tilde{\mathcal{G}}$ as long as one adheres to the constraint that $\tilde{\mathcal{G}}^{-1}$ and $\tilde{\mathcal{G}}^{-2}$ are well defined operators.  These constraints are necessary to ensure that all the standard CMB spectra are recoverable from the modified $E'$/$B'$ maps emerging from the modified operators on the Stokes parameter maps. Noting that that the standard spin raising $\eth^2$ and spin lowering $\bar{\eth}^2$ operators are special cases of these generalized operators, we presented a band limited representation of these operators.

\revisit{Discussion on results from polarized filaments filaments, compare and contrast with those in \cite{Zaldarriaga2001a}.}

There are several advantages and alternative analysis routes that emerge from employing these real space operators for analysis of CMB polarization maps. To begin with, one does not need to evaluate the spin harmonic functions as the real space operators only rely on computing the Euler angles which can be done on the fly and the $P_{\ell}^{2}$ functions need to be tabulated only once for analysis at some predetermined resolution. Note that this real space analysis method trivially generalizes to analyzing maps of arbitrary spin. Especially given the Green's function interpretation of these operators, their implementation is trivially parallelizable, since the $E/B$ contribution from the Stokes charge in each pixel can be evaluated independently. Also note that the evaluation of contribution from charges in the masked portion of the sky can be easily omitted, unlike in the harmonic analysis. Given the Green's function interpretation of the  real space operators and their locality, suggests the possibility of converting the Stokes parameter time ordered data to scalar $E/B$ time ordered data given the pointing model for the observations, which can then be subsequently converted to maps of $E/B$ modes using standard map making techniques. Since the real space operators offer the ability of tuning the locality of $E/B$ maps, this can be potentially exploited to reduce foreground contamination from distant parts of the sky. For instance the real space operators can be defined such that the locality of their radial kernels is varied on different portions of the sky, dictated by say the foreground morphology, resulting in some scalar $E'/B'$ maps.  While this idea seems promising, the usefulness of this implementation will depend on whether the standard $E/B$ mode spectra are recoverable from scalar maps derived in this fashion. Note that it would be fairly non-trivial to think of such implementations using conventional harmonic space methods.

Results demonstrating the equivalence of these real space methods to the conventional harmonic space methods and from explorations of some of the ideas discussed above will be presented in the second part of this series of papers.

>>>>

\section{Discussion}\label{sec:discussion}
In this work, we presented a \revisit{first} derivation of the real space operators on the sphere that transforms the Stokes vector to into a vector of scalars and vice versa. We also presented real space operators that directly decompose the full Stokes vector \vp{} into vector \vp{E} and \vp{B} that correspond to the respective scalar modes.  A vector-matrix notation which allows for concise book keeping of all the standard operations involved in the analysis of CMB polarization maps. This real-space analysis method trivially generalizes to maps of arbitrary spin.

These real space operators provide a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts.  Only the radial part depends on the band limit.  The azimuthal part of the operator is primarily responsible for handling the spin decomposition, while the radial weights determine the non-local dependence of the resulting fields on the original fields.
The radial part of the operator kernel is roughly self-similar.  For the kernel to fall to a fixed fraction of its maximum, the radial distance scales as $\ell_{\rm max}^{-1}$.  The angle where the kernel falls to one percent is approximately $\beta_0 = \ell_{0}/\ell_{\rm max}$ with $\ell_{0}=22$ (\fig{fig:rad_ker_decay}).

Our close study of the real space operators show that they can be expressed either as a Green's function or as a convolving beam function.  The interpretation  depends on whether the kernels are expressed in terms of the forward or inverse rotation Euler angles.  These functions are complex-valued and in many cases latitude-dependent.

The Green's function interpretation provides some useful insights into these operations. This interpretation allows us to think of ${}_{+2}X=Q+iU$ as some spin-2 charge which radiates out a complex spin-0 scalar field $E+iB$. The resulting complex scalar maps can be then understood as arising from superposition of the radiating fields emanating from all the spin charges on the sphere.  The $E/B$ mode maps are merely the real and imaginary parts of this field.  \revisit{A similar equation for real space $E/B$ operators was derived in \cite{Zaldarriaga2001a}, but treated the flat sky case and did not explicitly derive the radial kernel.}

Understanding the non-locality of the real space operators allowed us to generalize the real space operators % $\bar{O}'$ and its inverse ${\bar O}'^{-1}$, which
that transform between the spin-2 and spin-0 representations of the CMB polarization.
We showed how to modify and gereralize the construction of the scalar polarization fields and to control the radial kernels, specifying them with few restrictions.  These modifications to the radial kernel have the same effect as a smoothing operation on the scalar fields by a circularly symmetric beam.  Thus they are trivial to recover the standard $E/B$ power spectra in maps from the modified kernels. 
%We presented a method of systematically generalizing these operators by introducing the harmonic space operator $\tilde{\mathcal{G}}$ which in effect causes the radial function to vary from its default form. We argue that one can choose any arbitrary form for $\tilde{\mathcal{G}}$ as long as one adheres to the constraint that $\tilde{\mathcal{G}}^{-1}$ and $\tilde{\mathcal{G}}^{-2}$ are well defined operators.  These constraints are necessary to ensure that all the standard CMB spectra are recoverable from the modified $E'$/$B'$ maps emerging from the modified operators on the Stokes parameter maps.  
Noting that that the standard spin-raising ($\eth^2$) and spin-lowering ($\bar{\eth}^2$) operators are special cases of these generalized operators, we presented a band limited representation of these operators.

The toolbox of real-space operators gives us more intuition about the $E$- and $B$-mode structure of polarized filaments in the Milky Way gas and dust, potentially important sources of foregrounds for inflationary science.  We predict a characteristic $B$-mode pattern from filament sources.

%\revisit{Discussion on results from polarized filaments, compare and contrast with those in \cite{Zaldarriaga2001a}.}

Several alternative analysis routes emerge from considering  these real space operators.  No spin-harmonic transforms are necessary as the real space operators only rely on computing the Euler angles which can be done on the fly.  The radial functions (depending on $P_{\ell}^{2}$) need be tabulated only once at some determined resolution.  Especially given the Green's function interpretation of these operators, their implementation is trivially parallelizable, since the $E/B$ contribution from the Stokes charges in each pixel can be evaluated independently.  Although the scaling for the real space method (${\cal O}(N_{\rm pix}^2)$) is steeper than for the spherical harmonic transforms (${\cal O}(N_{\rm pix}^{3/2})$), the simple parallelization and the small prefactor for compact kernels at high-resolution / high-band-limit could make it faster in some contexts.  It can have a small memory footprint.  \comment{Does parallel S2hat have memory constraints at high l?}  The evaluation of contribution from charges in the masked portion of the sky can simply be omitted.  \comment{hybrid low-l / high-l schemes}

\revisit{The Green's function interpretation of the  real space operators and their locality, suggests the possibility of converting the Stokes parameter time ordered data to scalar $E/B$ time ordered data given the pointing model for the observations, which can then be subsequently converted to maps of $E/B$ modes using standard map making techniques.}  \comment{K: question whether this actually works.  What does E/B tod really mean?}

Since the real space operators let us tune the locality of $E/B$ maps, this can be potentially exploited to eliminate foreground contamination from distant parts of the sky. Such applications can be difficult to implement using conventional harmonic space methods.  For instance, the real space operators can be defined such that the locality of their radial kernels is varied on different portions of the sky, dictated by say the foreground morphology, resulting in some modified scalar $E'/B'$ maps.  While this idea seems interesting, the usefulness of this implementation will depend on whether the standard $E/B$ mode spectra are easily recoverable in this fashion.

Results that demonstrate the equivalence of these real space methods to the conventional harmonic space methods will be presented in the next paper in this series, along with explorations of some of the ideas discussed above.
